% ============================================================================
% Section: Data Sources and Data Generation
% ----------------------------------------------------------------------------
% Assumes: \usepackage[utf8]{inputenc}, \usepackage[T1]{fontenc}
%          \usepackage{booktabs}, \usepackage{array}
% (booktabs gives \toprule/\midrule/\bottomrule; array enables the p{} column
%  type used below for wrapped description columns; both are standard in
%  virtually all modern LaTeX distributions.)
% Adjust \label{...} keys if they collide with existing labels in the report.
%
% NOTE: this section documents Module 1 (Business Understanding & Data
% Generation) scope only. The dataset produced here was subsequently passed
% through a separate Data Cleaning stage (missing-value, duplicate,
% invalid-category, and outlier handling with a before/after report) that
% belongs in a different section of the report (typically the Data Cleaning
% module) and is therefore only cross-referenced, not duplicated, below.
% ============================================================================

\section{Data Sources and Data Generation}

This section documents, in full, the work performed to establish the data foundation of the fraud-detection pipeline: the base dataset and its empirically verified properties; the design, implementation, and independent validation of twelve synthetic contextual fields; and the resulting data dictionary. Emphasis is placed not only on the final artifacts but on the process by which they were produced and verified, since the reliability of every downstream modeling result depends on the rigor of this stage.

\subsection{Base Dataset}
\label{subsec:base-dataset}

\paragraph{Source.} The base transactional data is drawn from the \emph{Online Payments Fraud Detection Dataset}\footnote{\url{https://www.kaggle.com/datasets/rupakroy/online-payments-fraud-detection-dataset}} (commonly referred to as PaySim), a publicly available Kaggle dataset generated by the agent-based mobile-money simulator described by Lopez-Rojas, Elmir, and Axelsson (2016).\footnote{Lopez-Rojas, E. A., Elmir, A., \& Axelsson, S. (2016). PaySim: A financial mobile money simulator for fraud detection. In \emph{Proceedings of the 28th European Modeling and Simulation Symposium (EMSS)}.} The simulator models a population of customers and merchants performing financial transactions over a one-month period, with a small subset of transactions labeled as fraudulent to emulate account-takeover behavior (an attacker gaining unauthorized access to an account and subsequently transferring or withdrawing funds). This dataset was selected as the analytical base because it provides genuine transaction-level structure (balances, transaction types, and a verified fraud label) at a scale and class imbalance representative of real payment-fraud problems, rather than a synthetically balanced toy dataset that would understate the difficulty of the detection task.

\paragraph{Structure.} The dataset comprises \textbf{6,362,620 transaction records} across \textbf{11 columns}, summarized in Table~\ref{tab:paysim-columns}. The complete dataset was used throughout this work; no sampling or subsetting was applied at any stage, in order to preserve the true class-imbalance ratio and avoid introducing sampling bias into either the generation or the validation procedures described later in this section.

\begin{table}[htbp]
\centering
\caption{Base columns of the PaySim transactional dataset}
\label{tab:paysim-columns}
\footnotesize
\begin{tabular}{@{}llp{7.5cm}@{}}
\toprule
\textbf{Column} & \textbf{Type} & \textbf{Description} \\
\midrule
\texttt{step} & Integer & Elapsed simulation time, in hours, since the start of the simulation (range 1--743, i.e.\ approximately 31 days). \\
\texttt{type} & Categorical & Transaction type: \texttt{CASH\_IN}, \texttt{CASH\_OUT}, \texttt{DEBIT}, \texttt{PAYMENT}, or \texttt{TRANSFER}. \\
\texttt{amount} & Float & Transaction value, in the simulator's local currency unit. \\
\texttt{nameOrig} & String & Identifier of the account initiating the transaction (prefix \texttt{C}). \\
\texttt{oldbalanceOrg} & Float & Origin account balance immediately before the transaction. \\
\texttt{newbalanceOrig} & Float & Origin account balance immediately after the transaction. \\
\texttt{nameDest} & String & Identifier of the recipient account (prefix \texttt{C} for a customer, \texttt{M} for a merchant). \\
\texttt{oldbalanceDest} & Float & Destination account balance immediately before the transaction. \\
\texttt{newbalanceDest} & Float & Destination account balance immediately after the transaction. \\
\texttt{isFraud} & Binary & Ground-truth fraud label (1 = fraudulent, 0 = legitimate). \\
\texttt{isFlaggedFraud} & Binary & A naive rule-based flag built into the simulator (large-value transfers); present for only 16 of the 6,362,620 records and distinct from the true fraud label \texttt{isFraud}. \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Class imbalance.} The dataset exhibits the severe class imbalance typical of real-world fraud problems: only \textbf{8,213 of 6,362,620 transactions (0.1291\%)} are labeled fraudulent. This ratio was measured directly from the raw file (not taken from external documentation) and was subsequently used as a reference value to confirm, at every later processing stage, that no step in the pipeline silently altered the class distribution.

\paragraph{Empirical profiling methodology.} Before any generation logic was designed, the base dataset was profiled computationally rather than assumed to behave in an expected way. This profiling consisted of three concrete checks executed against the full 6,362,620-row file: (i) a cross-tabulation of the fraud label against transaction type, to determine which transaction types actually carry fraud risk; (ii) a cardinality check on the \texttt{nameOrig} identifier (count of unique values versus total row count), to determine whether multi-transaction customer histories exist within the data; and (iii) a range and gap check on the \texttt{step} column, to determine whether time-based features could be derived deterministically. Three findings from this profiling directly determined the design of the synthetic fields presented in Section~\ref{subsec:synthetic-data}:

\begin{enumerate}
    \item \textbf{Fraud is confined to two transaction types.} Fraudulent transactions occur exclusively within \texttt{TRANSFER} (4,097 of 532,909 transactions) and \texttt{CASH\_OUT} (4,116 of 2,237,500 transactions); the \texttt{PAYMENT} (2,151,495 transactions), \texttt{CASH\_IN} (1,399,284 transactions), and \texttt{DEBIT} (41,432 transactions) types contain \emph{zero} fraudulent records. This indicates that the fraud modeled in this dataset represents \emph{account-takeover} behavior (unauthorized transfer or cash-out from a compromised account) rather than point-of-sale or card-not-present payment fraud, a distinction that shaped the business narrative used to justify every conditional field described later.
    \item \textbf{Customer transaction history is negligible.} \texttt{nameOrig} takes 6,353,307 distinct values across the 6,362,620 rows (99.85\% of identifiers are unique to a single row); of these, 6,344,009 rows (99.71\% of the dataset) belong to an identifier that appears exactly once, while only 9,298 distinct identifiers recur (jointly accounting for the remaining 9,313 rows, since a handful of identifiers recur more than twice). Consequently, any feature requiring a persistent, multi-transaction customer profile (e.g.\ a rolling transaction-velocity counter) would be constructed on essentially non-existent history for the overwhelming majority of the population. All synthetic behavioral fields were therefore generated independently at the individual-transaction level rather than through an entity-level profile carried across multiple records.
    \item \textbf{The \texttt{step} field permits exact temporal derivation.} Because \texttt{step} encodes elapsed hours from a fixed simulation start with no gaps across its full 1--743 range, time-of-day features can be computed deterministically from this column rather than sampled stochastically, eliminating an unnecessary source of randomness (and thus of potential error) from the pipeline.
\end{enumerate}

This same profiling later served as the diagnostic basis for a subsequent, separately documented Data Cleaning stage, which confirmed the dataset to be structurally clean (no missing values, duplicate records, or invalid categorical values) and additionally flagged, without removing, three fraud-relevant anomalies (zero-amount transactions, statistical outliers in transaction amount, and a known balance-tracking characteristic of the simulator); that stage is out of scope for the present section and is reported separately.

\subsection{Synthetic Contextual Data}
\label{subsec:synthetic-data}

\paragraph{Motivation.} The base dataset contains only raw financial transaction fields and lacks the behavioral and e-commerce contextual signals ordinarily available to a production fraud-detection system: device fingerprinting, geolocation consistency between a transaction's IP address and the account's registered billing address, account tenure, payment-retry behavior, and temporal usage patterns. Twelve additional contextual fields were therefore generated in Python, using the \texttt{Faker} library in combination with custom, explicitly documented business logic, to bring the dataset closer to what a real payment platform would observe at the point of transaction.

\paragraph{Generation methodology and engineering approach.} All twelve fields were implemented as fully vectorized \texttt{numpy}/\texttt{pandas} operations applied to the complete 6,362,620-row dataset in a single pass, with no row-wise iteration and no sub-sampling, so that the procedure remains computationally tractable at full scale. All randomness was drawn from a single seeded generator (\texttt{numpy.random.Generator} with seed 42), so that re-running the identical pipeline against the identical input reproduces an identical output byte-for-byte --- a property that was explicitly verified rather than assumed. The implementation was organized into three focused modules with clearly separated responsibilities (a geographic-distance utility, the field-generation logic itself, and a leakage-validation and reporting module), and was developed test-first: every generator function is covered by dedicated unit tests that assert its statistical behavior (e.g.\ that a Bernoulli-conditioned field reproduces its specified base and fraud-conditional rates within tolerance, that a derived field is mathematically consistent with the columns it is derived from, and that boundary and degenerate cases are handled correctly). Fifty-two such tests were written and are run automatically; all fifty-two pass. Five design principles governed the generation process itself:

\begin{enumerate}
    \item \textbf{Row-level generation.} Given the negligible customer transaction history established in Section~\ref{subsec:base-dataset}, fields were generated independently per transaction rather than through a persistent customer profile carried across multiple records. This was a direct, empirically justified departure from the more conventional entity-profile design pattern used in commercial fraud systems, adopted only because the data itself does not support it.
    \item \textbf{Selective conditioning on the fraud label.} Only fields with a plausible behavioral connection to account-takeover fraud (use of an unrecognized device, a geographically inconsistent IP address, an unusually recent account, an atypical transaction hour, repeated failed payment attempts) were sampled conditionally on \texttt{isFraud}. Fields lacking such a rationale (browser, device type, device identifier, billing country) were deliberately generated independently of the label. This asymmetry is itself a design choice: a synthetic dataset in which \emph{every} field correlates with the target label would be unrealistic and would trivialize the classification task, so several fields were intentionally left uninformative.
    \item \textbf{Bounded effect sizes.} Every conditional parameter was capped at a modest effect size relative to its baseline --- roughly two to four times baseline for the Bernoulli probability and Poisson rate parameters, and a partial (rather than extreme) shift for the one continuous distributional parameter --- and documented as an explicit business assumption, rather than calibrated against any external ground truth, since no publicly available dataset documents these behavioral rates for the PaySim population. This bound was chosen specifically to avoid the field becoming a near-deterministic proxy for the label, which would constitute label leakage.
    \item \textbf{Deterministic derivation where possible.} Fields computable directly from existing columns --- the time-of-day features, and the geographic distance between the IP-inferred country and the billing country --- were derived by closed-form formula rather than sampled stochastically, both for accuracy and to guarantee internal consistency (e.g.\ the distance field is mathematically guaranteed to equal zero whenever the two country fields it depends on are equal).
    \item \textbf{Post-hoc, objective leakage validation.} After generation, every new field's univariate association with the fraud label was measured with a defined statistical procedure (described below) and checked against a fixed numeric threshold, rather than accepted on the basis of a subjective judgment that the field ``looked reasonable.''
\end{enumerate}

Table~\ref{tab:synthetic-fields} summarizes all twelve generated fields, their generation mechanism, and the specific parameters used.

\begin{table}[htbp]
\centering
\caption{Synthetic contextual fields added to the base dataset}
\label{tab:synthetic-fields}
\footnotesize
\begin{tabular}{@{}llp{7cm}@{}}
\toprule
\textbf{Field} & \textbf{Mechanism} & \textbf{Description} \\
\midrule
\texttt{hour\_of\_day} & Derived & Hour of day (0--23), computed as \((\texttt{step}-1) \bmod 24\); requires no assumption since it follows directly from the simulator's own clock. \\
\texttt{is\_night\_transaction} & Derived & Boolean flag for transactions occurring between 00:00 and 05:59, a conventional definition of ``night'' used in temporal fraud studies. \\
\texttt{customer\_account\_age\_days} & Conditional (log-normal) & Age of the customer account in days (bounded to [1, 3650]); median 400 days for legitimate transactions versus 275 days for fraudulent ones ($\sigma = 0.6$ on both), reflecting the assumption that compromised or mule accounts skew more recently created. \\
\texttt{device\_id} & Independent & Device identifier sampled uniformly from a fixed pool of 50,000 Faker-generated UUIDs; carries no signal on its own, so that the pool size and identity itself do not leak fraud information (the signal instead lives in \texttt{new\_device\_flag}). \\
\texttt{browser} & Independent & Browser family, sampled from a fixed market-share-like distribution (Chrome 55\%, Safari 20\%, Edge 12\%, Firefox 8\%, Other 5\%); deliberately independent of the fraud label. \\
\texttt{device\_type} & Independent & Device category (mobile 65\%, desktop 30\%, tablet 5\%); deliberately independent of the fraud label. \\
\texttt{new\_device\_flag} & Conditional (Bernoulli) & Whether the transaction originates from a device not previously associated with the account; base rate 4\%, rising to 12\% (3\(\times\) the baseline probability) for fraudulent transactions. \\
\texttt{billing\_country} & Independent & Customer's registered country, sampled from a fixed weighted distribution over 20 countries approximating a plausible customer-base geography. \\
\texttt{ip\_country} & Conditional (Bernoulli) & Country inferred from the transaction's IP address; matches \texttt{billing\_country} with probability 93\% for legitimate transactions and 80\% for fraudulent ones, and is guaranteed (by construction, via a modular offset over the fixed 20-country list) to differ from \texttt{billing\_country} on every non-matching row. \\
\texttt{ip\_billing\_distance\_km} & Derived & Great-circle (haversine) distance, in kilometers, between the \texttt{ip\_country} and \texttt{billing\_country} centroids, computed from a fixed country-centroid lookup table; exactly zero when the two countries coincide. \\
\texttt{shipping\_billing\_mismatch} & Conditional (Bernoulli) & Whether the transaction's associated address differs from the account's registered address; base rate 5\%, rising to 15\% for fraudulent transactions. \\
\texttt{failed\_payment\_attempts\_24h} & Conditional (Poisson) & Number of failed payment attempts in the preceding 24 hours; mean 0.15 for legitimate transactions, rising to 0.6 (a 4\(\times\) rate increase) for fraudulent ones. \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Leakage validation methodology.} A synthetic feature that correlates too strongly with the fraud label would trivialize the downstream classification task and misrepresent the analytical value of the generated context, so an explicit, quantitative validation step follows generation rather than a qualitative review alone. For continuous or Boolean fields, the univariate area under the ROC curve (AUC) of the field against \texttt{isFraud} was computed; because a feature that strongly predicts the \emph{absence} of fraud is exactly as informative (and exactly as concerning from a leakage standpoint) as one that strongly predicts its presence, the reported statistic is \(\max(\mathrm{AUC},\, 1-\mathrm{AUC})\), making the measure direction-agnostic. For categorical fields, an association statistic based on the chi-square test of independence between the field and the label was used; specifically, Cram\'er's V, but computed with Bergsma's (2013) small-sample bias correction\footnote{Bergsma, W. (2013). A bias-correction for Cram\'er's V and Tschuprow's T. \emph{Journal of the Korean Statistical Society}, 42(3), 323--328.} rather than the textbook formula. A field was flagged as failing the leakage check if its AUC reached or exceeded 0.75, or its corrected Cram\'er's V reached or exceeded 0.50; these thresholds were fixed in advance of running the check, not adjusted afterward to force a pass.

\paragraph{Iterative refinement: two issues identified and resolved.} The validation procedure was applied iteratively against the full dataset, and it was not treated as a formality --- it materially changed the generation logic on two separate occasions.

First, on the initial run against all 6,362,620 rows, \texttt{customer\_account\_age\_days} exceeded the AUC threshold (measured AUC \(=0.8753\)), indicating that the age gap between legitimate and fraudulent accounts, as originally parameterized (a fraud-conditional median of 150 days against a baseline of 400 days), separated the two classes too strongly to be considered a realistic behavioral signal rather than a near-deterministic proxy for the label. In response, the fraud-conditional median was narrowed from 150 to 275 days --- a deliberate reduction of the effect size rather than its outright removal, so that the field would continue to carry a genuine but non-decisive signal --- and the field was regenerated. Re-running the validation confirmed an AUC of 0.6689, below the threshold.

Second, a subsequent independent review of the validation code itself (rather than of its output) identified that the originally implemented Cram\'er's V formula, prior to the bias correction described above, could report a spuriously high, sample-size-dependent value for the high-cardinality \texttt{device\_id} field (50,000 distinct values): although this field is generated completely independently of the fraud label by construction, the uncorrected statistic was shown to be capable of reporting an association as high as approximately 0.48 on a moderately sized data subset, dangerously close to the 0.50 failure threshold, purely as an artifact of contingency-table sparsity rather than any genuine relationship. Because the corrected estimator explicitly removes this small-sample bias, it was substituted for the naive formula everywhere Cram\'er's V is computed, with no change to the generation logic itself; \texttt{device\_id}'s corrected association on the full dataset is exactly 0.0000.

Table~\ref{tab:leakage-results} reports the final validation results, measured on the complete 6,362,620-row dataset after both corrections; all twelve fields satisfy the leakage criteria.

\begin{table}[htbp]
\centering
\caption{Final leakage-validation results on the full dataset (6,362,620 rows)}
\label{tab:leakage-results}
\footnotesize
\begin{tabular}{@{}lccc@{}}
\toprule
\textbf{Field} & \textbf{Metric} & \textbf{Value} & \textbf{Status} \\
\midrule
\texttt{hour\_of\_day} & AUC & 0.6336 & PASS \\
\texttt{is\_night\_transaction} & AUC & 0.6217 & PASS \\
\texttt{customer\_account\_age\_days} & AUC & 0.6689 & PASS \\
\texttt{device\_id} & Corrected Cram\'er's V & 0.0000 & PASS \\
\texttt{browser} & Corrected Cram\'er's V & 0.0000 & PASS \\
\texttt{device\_type} & Corrected Cram\'er's V & 0.0004 & PASS \\
\texttt{new\_device\_flag} & AUC & 0.5419 & PASS \\
\texttt{billing\_country} & Corrected Cram\'er's V & 0.0000 & PASS \\
\texttt{ip\_country} & Corrected Cram\'er's V & 0.0049 & PASS \\
\texttt{ip\_billing\_distance\_km} & AUC & 0.5651 & PASS \\
\texttt{shipping\_billing\_mismatch} & AUC & 0.5495 & PASS \\
\texttt{failed\_payment\_attempts\_24h} & AUC & 0.6589 & PASS \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Effect on class distribution.} Generating the twelve fields did not alter the underlying class distribution: the resulting dataset retains all 6,362,620 records and the original 0.1291\% fraud rate. Class-imbalance remediation (e.g.\ oversampling, undersampling, or class weighting) was deliberately deferred to the feature-engineering and modeling stage, so that this stage's sole responsibility is enriching the data, not altering its label distribution. The enriched 23-column dataset was written out as the input to the subsequent data-cleaning stage, together with a smaller stratified sample (drawn to preserve the 0.1291\% fraud rate) for quick manual inspection.

\paragraph{Limitations.} The conditional generation parameters reported in Table~\ref{tab:synthetic-fields} are explicit business assumptions rather than values calibrated against an external ground truth, since no publicly available dataset documents these behavioral rates for the PaySim population; this is an inherent limitation of any synthetic-augmentation approach applied to a dataset that was not originally collected with this contextual information. Because the base dataset's fraud label represents account-takeover behavior confined to the \texttt{TRANSFER} and \texttt{CASH\_OUT} transaction types (Section~\ref{subsec:base-dataset}), one field --- \texttt{shipping\_billing\_mismatch} --- was reframed from its literal e-commerce checkout interpretation to an account-takeover interpretation (a discrepancy between the transaction's associated address and the account's registered address), in order to remain consistent with the base dataset's actual fraud mechanism rather than importing an inconsistent narrative. Finally, all twelve fields were generated independently at the transaction level (Design Principle 1); should a future extension of this dataset introduce genuine repeat-customer histories, entity-level behavioral features (e.g.\ transaction velocity per account over a rolling time window) could be constructed in addition to the row-level fields presented here.

\subsection{Data Dictionary}
\label{subsec:data-dictionary}

\paragraph{Content and structure.} In accordance with the requirement that every synthetically generated field be formally documented, a structured data dictionary was produced covering all twelve fields listed in Table~\ref{tab:synthetic-fields}. For each field, the dictionary records eight attributes: the column name; the data type; the unit; the valid range; the generation mechanism (derived, independently sampled, or conditional on the fraud label); the exact generation formula or distribution, written out explicitly rather than described in prose; the underlying business assumption motivating the field and its parameterization; and the measured univariate association with \texttt{isFraud} reported in Table~\ref{tab:leakage-results}, labeled with its metric type (AUC or Cram\'er's V) so that the two different scales are not confused with one another.

\paragraph{Automated generation.} Rather than being transcribed by hand, the data dictionary is generated programmatically by the same validation module that performs the leakage checks described in Section~\ref{subsec:synthetic-data}: the module first computes each field's association statistic and then renders the dictionary as a formatted table directly from that same in-memory result. This design guarantees that the documented specification and the executed generation code cannot drift out of sync --- any future change to a generation parameter is automatically reflected in the published dictionary the next time the pipeline is run, and every reported association value is a direct, reproducible measurement on the current dataset rather than a manually maintained (and potentially stale) estimate. This process itself caught one documentation defect during development: an early version of the rendered dictionary reported the numeric association value without indicating which metric it belonged to, which was corrected so that every value is now explicitly labeled (e.g.\ ``0.6336 (AUC)'' versus ``0.0049 (Cram\'er's V)'').

\paragraph{Format and scope.} The dictionary is published in Markdown table format, satisfying the required deliverable format (Markdown or Excel), and is available as a standalone artifact alongside the generation code. Table~\ref{tab:data-dict-excerpt} illustrates its structure with three representative entries; the full dictionary contains all twelve fields.

\begin{table}[htbp]
\centering
\caption{Excerpt of the generated data dictionary (illustrative entries)}
\label{tab:data-dict-excerpt}
\footnotesize
\begin{tabular}{@{}lp{2.2cm}p{2.2cm}p{6cm}@{}}
\toprule
\textbf{Field} & \textbf{Type / Range} & \textbf{Generation} & \textbf{Business assumption} \\
\midrule
\texttt{new\_device\_flag} & Boolean & Conditional (Bernoulli, $p=0.04\!\to\!0.12$) & Account-takeover fraud is plausibly more often initiated from an unrecognized device, though not certainly so, since sophisticated attackers may spoof a known device fingerprint. \\
\texttt{ip\_billing\_distance\_km} & Float, [0, $\sim$20{,}000] km & Derived from \texttt{ip\_country} / \texttt{billing\_country} & Computed directly from a fixed country-centroid lookup table to guarantee internal consistency with the country-mismatch fields. \\
\texttt{failed\_payment\_attempts\_24h} & Integer, [0, $\sim$5] & Conditional (Poisson, $\lambda=0.15\!\to\!0.6$) & Attackers are assumed to make repeated failed attempts (e.g.\ credential or card guessing) before a successful transaction. \\
\bottomrule
\end{tabular}
\end{table}
